---
title: "Part 2"
author: "Ashton Koerner"
date: "2023-11-28"
output: pdf_document
---

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
library(dplyr)
library(MASS)
```

## Part 2
### 3.1
```{r}
# reading data and changing categorical entries to numeric entries
data = read.csv('iris_data.data')
colnames(data) = c("sepal.length","sepal.width","petal.length", "petal.width","class")
data$Numeric_class = as.numeric(factor(data$class, levels = unique(data$class)))
data = data[,-5]
names(data)[names(data) == 'Numeric_class'] <- 'species'
# 1 = Iris-setosa
# 2 = Iris-versicolor
# 3 = Iris-virginica


# Orthonormal basis for subspace U using PCA
U = prcomp(data)$rotation

# Function computing the novelty of an instance x with respect to subspace U
novelty = function(x, U) {
  projection = x %*% U
  novelty = sqrt(sum((x - projection %*% t(U))^2))
  return(novelty)
}

# Applying novelty function to each row of data
novelties = apply(data, 1, function(x) novelty(x, U))

plot(novelties, type = "l", xlab = "Instance",
     ylab = "Novelty", main = "Novelty of Instances")
summary(novelties)
IQR = 2.809e-15 - 1.814e-15
Upper = 2.809e-15 + 1.5*IQR
Lower = 1.814e-15 - 1.5*IQR
# Identifying instances that stand out
outliers = which(novelties > Upper)
outliers = which(novelties < Lower)
cat("Instances that stand out:", outliers, "\n")
```

Outlier detection aims to find anomalies within a dataset, while novelty detection focuses on identifying instances that are novel or different from what the model has seen during training.

No instances that stand out mean in this context that no one iris plant is more interesting than the rest.

### 3.2 Classification
```{r}
# Fisher Discriminant Analysis
iris_binary = subset(iris, Species %in% c("setosa", "versicolor"))
iris_binary$Species = as.factor(iris_binary$Species)
lda_model = lda(Species ~ Sepal.Length + Sepal.Width, data = iris)
lda_model
```

The mean Sepal Length for setosa is 5.006, and the mean Sepal Width is 3.428.
The mean Sepal Length for versicolor is 5.936, and the mean Sepal Width is 2.770.
The mean Sepal Length for virginica  is 6.588, and the mean Sepal Width is 2.974.

In LD1, Sepal.Length has a coefficient of approximately -2.14, and Sepal.Width has a coefficient of approximately 2.77. These coefficients indicate the weights assigned to the predictors in creating LD1 and LD2.
The coefficients of linear discriminants (LD1 and LD2) indicate the importance of each predictor in separating the groups. Negative and positive coefficients contribute to the direction of separation.
D1 captures approximately 96.28% of the total variance, while LD2 captures 3.72%.
The proportion of trace indicates the importance of LD1 and LD2 in capturing the overall variability.

Overall, the results suggest that LD1, which is a linear combination of Sepal Length and Sepal Width, is the primary factor responsible for the separation between the groups. 

## Part 3
```{r}
iris_features <- iris[, 1:4]
pca_result <- prcomp(iris_features, scale. = TRUE)
pc1 <- pca_result$x[, 1]
pc2 <- pca_result$x[, 2]
plot(pc1, pc2, col = iris$Species, pch = 19, main = "PCA of Iris Data", xlab = "P", ylab = "PC2")
legend("topright", legend = levels(iris$Species), col = 1:3, pch = 19, title = "Species")
```

Setosa species is clustered away from the other two species along PC1. Versicolor and virginica species show some overlap, indicating that these species share certain similarities in their measurements. 
A general trend is that versicolor tends to have lower values along PC2, while virginica tends to have higher values.
